Extended Data

### 2022-03-01 doc/manuscript/Dow_manuscript_Nature.Rmd
rearrange data cleaning section

The raw dendrometer band data were manually inspected before analysis. 
We screened the data for three classes of errors. 
First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and removed.
Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck on the tree bark and then released. 
In these cases, the full annual record for the tree was removed.
Third, data points that deviated substantially from normal growth patterns, but for unknown causes, were removed. 
If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.

@@

After fitting the growth model, we removed tree-years with poor fits. 
Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the logical range. 
Modeled fits for tree-years were removed under five conditions:
<!--rearrange these in order of how data were removed (3 & 4 come first)-->
(1) ~~single day growth rates were~~ $g_{max}$ was <!--confirm this change--> $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101); 
(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253); 
(4) tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, HF = 8).
At both sites the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
This process was repeated using 2 and 3 standard deviations and the results of analyses were not qualitatively different.

>>>>

The raw dendrometer band data were manually inspected before analysis. 
We screened the data for three classes of errors. 
First, when a measurement was drastically different from previous and following measurements, it was assumed to be a human error and removed.
Second, when measurements remained essentially unchanged for several readings, followed by a sudden jump then return to a normal growth pattern, this was assumed to be a case where the band was stuck on the tree bark and then released. 
In these cases, the full annual record for the tree was removed.
Third, data points that deviated substantially from normal growth patterns, but for unknown causes, were removed. 
If a majority of the data points fell into this class within a tree-year, the entire year was removed from the analysis.
We also removed tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.05$ mm; SCBI = 26, Harvard Forest = 253), and tree-years where first intraannual measurement was later than the first spring survey (trees that were missed in the initial census; SCBI = 22, HF = 8).
These were removed because they reduce the reliability of predicted growth in the modeled curves.

@@

After fitting the growth model, we removed tree-years with poor fits. 
Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the logical range. 
Modeled fits for tree-years were removed under five conditions:
<!--rearrange these in order of how data were removed (3 & 4 come first)-->
(1) ~~single day growth rates were~~ $g_{max}$ was <!--confirm this change--> $\ge 2.5$ standard deviations away from the mean for each xylem architecture (SCBI = 3, Harvard Forest = 11); 
(2) timing variables ($DOY_{ip}$, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$) were $\ge 2.5$ standard deviations away from the means for their xylem architecture group, year, and site (SCBI = 74, Harvard Forest = 101).
At both sites the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others.
This process was repeated using 2 and 3 standard deviations and the results of analyses were not qualitatively different.
